---
title: "Causal inference"
subtitle: "in large N analysis"
author: "Austin Hart"
institute: "American University"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, rladies, rladies-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  message=FALSE, warning=FALSE, echo = FALSE, 
  fig.width=3, fig.height=2, fig.align = 'center'
)
```

```{r results='hide', message=FALSE, warning=FALSE, eval=TRUE}
library(tidyverse)
library(magrittr)
```

## Statistics and causal inference

- **CAUSAL INFERENCE**: The attribution of an effect to a cause based on the evidence presented.

- **STATISTICS**: Science of analyzing data collected over a large number of observations.

- **INTERNAL VALIDITY**: Extent to which evidence presented supports a causal claim.  
i.e. can eliminate alternative explanations for a finding. 


---
class: inverse, center, middle

# Criteria for causality

---

# Hurdles to internal validity

Given the argument that A is causing B:

- Is there **evidence that A and B covary**?

- Is there a **credible causal mechanism** linking A to B?

- Can you **rule out the reverse claim**, that B is causing A?

- Can you **rule out confounding variables**, or does A share a common cause with B that makes the relationship spurious?


---

# Hurdle 1: covariation

Is there statistical evidence that the two factors, A and B, covary?

- **WHO CARES?** Without it, all you have is an assertion. 
- **WHAT TO LOOK FOR:** Evidence that A and B 'move together'  

  > "Risk of heart attack among vegetarians was 13% lower than among meat eaters"
    
- **RAISING THE ALARM:**  

  > "The study fails to show that A and B are related at all. There is no empirical basis for inferring that A is causing B."  


---

# Hurdle 2: causal mechanism

Is there a plausible reason why exposure to A would lead to a response in B?

- **WHO CARES?** Without it, covariation b/w A and B may be coincidental.
- **WHAT TO LOOK FOR:** Simple explanation

  > "Vegetarian diet is lower in saturated fats which tend to create the blockages that lead to heart attacks"
  
- **RAISING THE ALARM:** 

  > "Without a plausible explanation as to how A would cause B, we cannot rule out the possibility that the observed association is purely coincidental."


---

# Hurdle 3: reverse causality

Given the claim that A is causing B, can you rule out the opposite interpretation: that B is causing A?

- **WHO CARES?** If not, you cannot distinguish cause from effect.
- **WHAT TO LOOK FOR:** Evidence that A *precedes* change in B.

  > "Those switching to vegetarian diet saw a 10% reduction in all cause deaths in the first year."
  
- **RAISING THE ALARM:**

  > "We cannot rule out the reverse causal claim that B is instead the cause of A."  

---

# Hurdle 4: confounding/selection bias

Can you **rule out confounding variables**, or do exposure (A) and outcome (B) share a common cause?

- **WHO CARES?** Violates "independence" assumption; renders association b/w A and B spurious
- **WHAT TO LOOK FOR:** Clear evidence of controlling for, conditioning on, factors that differentiate groups of A.  

  > "The analysis controls for respondent age and health history."
  
- **RAISING THE ALARM** 

  > "The study fails to control for a common cause of diet and heart health: welath. So the relationship may be spurious, reflecting the fact that wealthy adults are more likely to be vegetarian and, critically, more likely to have healthier hearts independent of their diet."  

---

# Applying the criteria in practice

- **IDENTIFY THE CAUSAL CLAIM**

  > "Eating a plant-based diet reduces risk of heart disease."
  
- **WORK THROUGH EACH HURDLE** with reference to the evidence presented.
- **IDENTIFY AND EXPLAIN ANY VIOLATIONS**

  > "The study fails to control for a common cause of diet and heart health: welath. So the relationship may be spurious, reflecting the fact that wealthy adults are more likely to be vegetarian and, critically, more likely to have healthier hearts independent of their diet."



---
class: inverse, center, middle

# Designs for Statistical Analysis


---

# Two modes of large-N study

- **EXPERIMENTAL (randomized control trial, RCT)**
  - Experimenter randomly assigns units to treatment
  - Experimenter controls exposure to treatment  

- **OBSERVATIONAL (epidemiological)**
  - Observer records scores on exposure variable
  - No control over exposure or environment


---
class: center, middle

# Experimental design

---

# Experiments (RCTs)

- **Process**
  - Gather subjects; RANDOMIZE into treatment groups
  - Administer treatment in controlled environment
  - Record outcome; compare across groups  
  
- **Why it works**
  - Randomization! Makes treatment groups identical in expectation
  - Control! Only the value of treatment variable differentiates groups
  - Eliminates confounds + rules out reverse causality


---

# Experimental example

> *What is the effect of plant-based diet on cholesterol?* 
>
> In a double-blind experiment, we randomly assigned 900 institutionalized adult subjects to either a plant-based or omnivorous diet for twelve months. 


```{r plot1,fig.align='center', out.width="50%",dev='svg'}
  df = tribble(~x, ~y,
               "Plant-based", 170,
               "Omnivorous", 235)

  ggplot(df, aes(x,y)) + 
    geom_col(fill = 'cornflowerblue',color = 'gray15') +
    scale_y_continuous(
      breaks = seq(0,250,50),
    ) +
    coord_cartesian(ylim = c(0,251)) +
    labs(
      x = "Assigned diet",
      y = "LDL (mg/dL)",
      title = "Cholesterol by diet"
    ) +
    theme_bw()
```

---
class: center, middle

# Observational designs


---

# Observing (not treating)

- **Process**
  - Identify sample
  - Record value on exposure and outcome variables
  - Comparison across groups, conditional on confounds  
  
- **When it works**
  - *"Strong exogeneity"*:  A way to make the treatment arguably or statistically independent
  - *Careful supplemental analyses*: Think "Law and Data" approach
  

---

# Observational remake

> *What is the effect of plant-based diet on cholesterol?* 
>
> We surveyed 900 adults about their dietary habits and cholesterol level.  


```{r plot2,fig.align='center', out.width="50%",dev='svg'}
  df = tribble(~x, ~y,
               "Vegetarian", 170,
               "Meat-eater", 235)

  ggplot(df, aes(x,y)) + 
    geom_col(fill = 'firebrick3',color = 'gray15') +
    scale_y_continuous(
      breaks = seq(0,250,50),
    ) +
    coord_cartesian(ylim = c(0,251)) +
    labs(
      x = "Reported diet",
      y = "LDL (mg/dL)",
      title = "Cholesterol by diet"
    ) +
    theme_bw()
```

---

# Comparing designs: what gives?

- **With random assignment + control**
  - No systematic differences b/w subjects across groups.
  - Difference in LDL attributable to only difference: diet  
    &zwnj;  
  
- **With self-selection**
  - vegetarians are younger, wealthier, more insured, etc.
  - Difference in LDL attributable to any/all of these things  
    &zwnj;  

Bottom line: the baseline for internal validity is MUCH higher in with an experimental design.


---
class: inverse, center, middle

# Examples for the week

---

# 1. Cost-sharing and malaria prevention

Cohen and Dupas (2010) "Free distribution or cost-sharing? Evidence from a randomized malaria prevention experiment."

- Causal claim?

- Study design?

- What do we learn from this?

---

# 2. Water fetching and child health

Pickering and Davis (2012) "Freshwater Availability and Water Fetching Distance Affect Child Health in Sub-Saharan Africa."

- Causal claim?

- Study design?

- How do they make fetching distance "exogenous"?

- What do we learn from this?


---
class: center, middle

# Problem Set 1

PS1 is due Monday, February 22.

[Find PS1 here](https://docs.google.com/document/d/1nUE3LVIaPmSoFHvMUBoPJyXfrPST3gmlGw_Oxuv4WDc/edit?usp=sharing)

Submit your completed work on Blackboard.
